---
title: "Quick start - How to run CoordinateCleaner"
output: html_document
---

CoordinateCleaner implements three wrapper functions to identify errors common in biological collections. The `clean_coordinates` function compares coordinates to geographic gazetters to identify potentially problematic records, the `clean_dataset` function uses dataset properties to test if rasterized collection might be an issue and `clean_fossils` includes a temporal dimension to clean fossils. 

***Please find a detailed tutorial on how to clean occurrence records (e.g. from GBIF) [here](https://azizka.github.io/CoordinateCleaner/articles/Cleaning_GBIF_data_with_CoordinateCleaner.html) and how to clean fossil data (e.g. from PBDB) [here](https://azizka.github.io/CoordinateCleaner/articles/Cleaning_PBDB_fossils_with_CoordinateCleaner.html).***

You can run Coordinate cleaner with few lines of R code:

```{r, collapse = T, tidy = T}
library(CoordinateCleaner)

#create example dataset
exmpl <- data.frame(species = sample(letters, size = 250, replace = TRUE),
                    decimallongitude = runif(250, min = 42, max = 51),
                    decimallatitude = runif(250, min = -26, max = -11),
                    dataset = "clean")

#run wrapper function
cord.flags <- clean_coordinates(x = exmpl)

#summarize and visualize results
summary(cord.flags)
plot(cord.flags)

#run dataset-level cleaning
ds.flags <- clean_dataset(exmpl)

```

Alternatively, you can run the individual functions in a pipe compatible way:
```{r, collapse = T, tidy = T}
library(tidyverse)

cleaned <-  exmpl%>%
  cc_val()%>%
  cc_cap()%>%
  cc_cen()%>%
  cc_dupl()%>%
  cc_equ()%>%
  cc_gbif()%>%
  cc_inst()%>%
  cc_outl()%>%
  cc_sea()%>%
  cc_zero()%>%
  cd_ddmm()%>%
  cd_round()
```
